\sysfull

### 2017-11-13 paper/background.tex

Energy harvesting devices operate using energy extracted from sources such as RF and solar energy. These devices elide tethered power or a battery, instead collecting energy into a capacitor, operating when sufficient energy accumulates, and upon depleting the energy, turning off and recharging.
%
%Batteryless operation has a number of important advantages, making intermittent computing an important research domain. Supplying power to billions~\cite{gartner_iot} of embedded computers using batteries is not sustainable. The European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Batteries are an environmental risk, are fragile, are limited in their number of charge/discharge cycles, and may require costly physical maintenance that is difficult or impossible deployed (e.g. in space~\cite{kicksat}). By contrast, super-capacitors are durable, promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}. The main limitation in moving to capacitor-based energy storage is that capacitor energy density---and consequently operating discharge time---is orders of magnitude less than a battery. 
%
%Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015} and digestible~\cite{nadeau_naturebio_2017} sensors.
%
Many platforms enable intermittent, battery-less energy harvesting-based computation. For instance, computational RFIDs---open-source TI MSP430-based~\cite{wolverine} WISP~\cite{wisp5} (with its variants such as WISPCam~\cite{naderiparizi_rfid_2015}, NFC-WISP~\cite{zhao_rfid_2015} or NeuralWISP~\cite{holleman_biocas_2008}), Moo~\cite{moo}, and commercial ones such as~\cite{medusa_farsens_2017}. Other intermittently-powered platforms include ambient backscatter tag~\cite{liu_sigcomm_2013,parks_sigcomm_2014} or battery-less phone~\cite{talla_imwut_2017}.

@@

\subsection{Energy Harvesting and Intermittent Computation}
\label{sec:background_consistency}

%Energy-harvesting devices execute software according to the {\em intermittent execution model}. Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle.
%
Software running on an energy-harvesting device executes {\em intermittently} because buffered energy is only sometimes available. An intermittent execution is composed of operating periods interspersed with power failures~\cite{dino,chain,alpaca,ratchet}. The frequency of failures depends on the size of the device's energy storage buffer: a larger buffer allows longer operating periods. Energy-harvesters provide input power orders of magnitude less than operating power, making recharging negligible during operation.

@@

\begin{figure}
	\centering
	%\subfloat[Simplified C code snippet of a CRC calculation from~\cite{hicks_mibench2_2016}: per-byte message division by a polynomial; \texttt{NV} denotes non-volatile variable declaration]{\includegraphics[width=\columnwidth]{figures/crc_example}\label{fig:crc_example}}\\
	%\subfloat[Consecutive execution steps of the loop body in the snippet above: non-volatile checkpointing did not guarantee data consistency as data has been manipulated (line 9) with stale reminder (line 3)]{\includegraphics[width=\columnwidth]{figures/crc_example_war}\label{fig:crc_example_war}}
	\caption{Code example demonstrating effect of write after read on volatile memory checkpointing.}
	\label{fig:code_demo_incosistency}
\end{figure}

Checkpoints of volatile state preserve intermittent progress, but do not ensure data consistency~\cite{dino,chain,ratchet}. Data may become inconsistent if an attempt to execute some computation {\em writes} to a non-volatile variable, then power fails, then a second attempt to re-execute the same computation incorrectly {\em reads} the value written in the first attempt, rather than the variable's original value. The situation occurs when code includes a write after read (WAR) dependence between operations that manipulate non-volatile variables~\cite{ratchet,dino,alpaca}.

Figure~\ref{fig:code_demo_incosistency} illustrates how state can become inconsistent in an intermittent execution using the cyclic redundancy check (CRC) code from MIBench~\cite{hicks_mibench2_2016}. The code computes the CRC for an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder $\texttt{Rmnd}$.

Despite checkpointing at line 3 in Fig.~\ref{fig:crc_example}, the code may compute  $\texttt{data}$ incorrectly because of the read of $\texttt{Rmnd}$ on line 4 and the write of $\texttt{Rmnd}$ on line 5. Figure~\ref{fig:crc_example_war} shows an intermittent execution. The second iteration writes $\texttt{Rmnd}$, but power fails before the checkpoint. After restarting, line 9 reads the updated value of $\texttt{Rmnd}$, producing an incorrect $\texttt{data}$ value.
%
Checkpoint-based systems risk violating memory consistency~\cite{dino}. To maintain consistency, prior systems~\cite{dino,ratchet} {\em version} a subset of non-volatile data with the checkpoint. Task-based systems~\cite{chain,alpaca}, which we focus on, ensure consistency using programmer, compiler, and runtime support.

\subsection{Task-based Intermittent Computation}
\label{section:background_task_computing}

@@

\noindent \textbf{Memory Consistency Enforcement Overhead:} Intermittent execution models incur overheads to checkpoint data~\cite{dino,ratchet,quickrecall,mementos}, manage channels~\cite{chain}, or privatize and commit data~\cite{alpaca}. Table~\ref{table:chechpoint_comparison} compares overheads for recent intermittent execution models (cf. \cite[Sec. 2.4]{alpaca}.) The mechanism responsible for overhead in each model varies, but the bulk of overhead in all models is a manipulating non-volatile memory. Checkpointing moves data to and from non-volatile memory, channel accesses manipulate non-volatile data~\cite{chain}, privatization copies from and commits to non-volatile memory~\cite{alpaca}, and idempotence solutions~\cite{ratchet} use {\em only} non-volatile memory.

\begin{figure}
	\centering
	%\subfloat[Energy cost of \emph{write} operation]{\includegraphics[width=0.49\columnwidth]{figures/fram_write.eps}}
	%\subfloat[Energy cost of \emph{read} operation]{\includegraphics[width=0.49\columnwidth]{figures/fram_read.eps}}
	\caption{Cost of accessing volatile (SRAM)/non-volatile (FRAM) memory during write (left) and read (right) operation.}
	\label{fig:framEnergy}
\end{figure}

To understand these non-volatile memory overheads, we measured the energy consumption of volatile (SRAM) and non-volatile (FRAM) memory operations for the MSP430FR5969~\cite{msp430datasheet} MCU. We used EDB~\cite{edb} to monitor voltage drop on a capacitor powering the device across 1600 read and write accesses to each memory type, randomizing to avoid caching. Figure~\ref{fig:framEnergy} summarizes the data, showing that a non-volatile memory access consumes around 1.5 times the energy of a volatile access in this device. In an intermittent device, energy cost corresponds to decreased run time between power failures. The cost of non-volatile memory motivates \sys, which virtualizes memory to primarily use SRAM.

@@

\noindent \textbf{Fixed-size Tasks are Inflexible and Inefficient:} If a
programmer writes a task that consumes {\em more} energy than the device can
buffer, the task will never complete using buffered energy. {\em preventing
forward progress} by repeatedly re-executing. If a programmer writes tasks that
consume far {\em less} energy than a device can buffer, the system may operate
{\em inefficiently}. When a task and its successor both complete, the two tasks
were interleaved by a task boundary that preserves progress and state (i.e.,
checkpoint or commit). However, absent a power failure, the work to preserve
state was {\em unnecessary}, incurring overhead on the tasks' executions.
Avoiding excessively costly, non-terminating tasks and short, high-overhead
tasks is a programming challenge, given fixed hardware with a fixed energy
buffering capacity. Buffer sizes from prior work vary widely from $20\mu F$~\cite{rodriguez_tbcs_2015}
to $0.1F$~\cite{moo}.
Sizing tasks complicates 
{\em porting} code from one device to another. An excessively costly task 
on a device with a small energy buffer may be a relatively short,
task on a device with a larger buffer. The
key problem is that using existing systems, the programmer statically sizes tasks for a fixed energy buffer. \sys's {\em task coalescing}, described
in Section~\ref{sec:task_coalescing}, is motivated by these observations.

\subsection{Hardware Assumptions}
\label{sec:background_hardware}

\sys is designed for the demands of existing and future energy-harvesting platforms based around general purpose, commodity computing components~\cite{wisp,msp430datasheet}. We assume a device with a memory system that has fast, byte-addressable volatile and non-volatile memory; in particular, our target platform~\cite{wisp} is equipped with a mixture of SRAM and FRAM. Our implementation leverages hardware support for fast, bulk-copying between memories via DMA~\cite{msp430datasheet}. We do not require a particular non-volatile memory technology, nor do we require architectural additions commodity processors~\cite{su_date_2017,ratchet,quickrecall,nvp}. \sys supports I/O behavior similar to~\cite{alpaca,chain}, allowing safe, synchronous I/O and unsafe, asynchronous I/O.

>>>>

Energy harvesting devices operate using energy extracted from sources such as radio frequency transmissions and solar energy. These devices elide tethered power or a battery, instead collecting energy into a capacitor, operating when sufficient energy accumulates, and upon depleting the energy, turning off and recharging.
%
%Batteryless operation has a number of important advantages, making intermittent computing an important research domain. Supplying power to billions~\cite{gartner_iot} of embedded computers using batteries is not sustainable. The European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Batteries are an environmental risk, are fragile, are limited in their number of charge/discharge cycles, and may require costly physical maintenance that is difficult or impossible deployed (e.g. in space~\cite{kicksat}). By contrast, super-capacitors are durable, promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}. The main limitation in moving to capacitor-based energy storage is that capacitor energy density---and consequently operating discharge time---is orders of magnitude less than a battery. 
%
%Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015} and digestible~\cite{nadeau_naturebio_2017} sensors.
%
Many platforms enable intermittent, battery-less energy harvesting-based computation. For instance, computational RFIDs---open-source TI MSP430-based~\cite{wolverine} WISP~\cite{wisp5} (with its variants such as WISPCam~\cite{naderiparizi_rfid_2015}, NFC-WISP~\cite{zhao_rfid_2015} or NeuralWISP~\cite{holleman_biocas_2008}), Moo~\cite{moo}, and commercial ones such as~\cite{medusa_farsens_2017}. Other intermittently-powered platforms include ambient backscatter tag~\cite{liu_sigcomm_2013,parks_sigcomm_2014} or battery-less phone~\cite{talla_imwut_2017}.

@@

\textbf{Hardware Assumptions.} \sys is designed for the demands of existing and future energy-harvesting platforms based around general purpose, commodity computing components~\cite{wisp,msp430datasheet}. We assume a device with a memory system that has fast, byte-addressable volatile and non-volatile memory; in particular, our target platform, WISP~\cite{wisp}, is equipped with a mixture of SRAM and FRAM. Our implementation leverages hardware support for fast, bulk-copying between memories via DMA~\cite{msp430datasheet}. We do not require a particular non-volatile memory technology, nor do we require architectural additions commodity processors~\cite{su_date_2017,ratchet,quickrecall,nvp}. \sys supports I/O behaviour similar to~\cite{alpaca,chain}, allowing safe, synchronous I/O and unsafe, asynchronous I/O.

\subsection{Intermittent Execution}
\label{sec:background_consistency}

%Energy-harvesting devices execute software according to the {\em intermittent execution model}. Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle.
%
Software running on an energy-harvesting device executes {\em intermittently} because buffered energy is only available \emph{sometimes}. An intermittent execution is composed of operating periods interspersed with power failures~\cite{dino,chain,alpaca,ratchet}. The frequency of failures depends on the size of the device's energy storage buffer: a larger buffer allows longer operating periods. Energy-harvesters provide input power orders of magnitude less than operating power, making recharging negligible during operation.

@@

\begin{figure}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example}
		\caption{Simplified C code snippet of a CRC calculation from~\cite{hicks_mibench2_2016}: per-byte message division by a polynomial; \texttt{NV} denotes non-volatile variable declaration.}\label{fig:crc_example}
	\end{subfigure}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example_war}
		\caption{Consecutive execution steps of the loop body in the snippet above: non-volatile checkpointing did not guarantee data consistency as data has been manipulated (line 9) with stale reminder (line 3)}\label{fig:crc_example_war}
    \end{subfigure}
	\caption{Code example demonstrating effect of write after read on volatile memory checkpointing.}\label{fig:code_demo_incosistency}
\end{figure}

%\begin{figure}
%	\centering
%	\subfloat[Simplified C code snippet of a CRC calculation from~\cite{hicks_mibench2_2016}: per-byte message division by a polynomial; \texttt{NV} denotes non-volatile variable declaration]{\includegraphics[width=\columnwidth]{figures/crc_example}\label{fig:crc_example}}\\
%	\subfloat[Consecutive execution steps of the loop body in the snippet above: non-volatile checkpointing did not guarantee data consistency as data has been manipulated (line 9) with stale reminder (line 3)]{\includegraphics[width=\columnwidth]{figures/crc_example_war}\label{fig:crc_example_war}}
%	\caption{Code example demonstrating effect of write after read on volatile memory checkpointing.}
%	\label{fig:code_demo_incosistency}
%\end{figure}

Checkpoints of volatile state preserve intermittent progress, but do not ensure data consistency~\cite{dino,chain,ratchet}. Data may become inconsistent if an attempt to execute some computation {\em writes} to a non-volatile variable, then power fails, then a second attempt to re-execute the same computation incorrectly {\em reads} the value written in the first attempt, rather than the variable's original value. The situation occurs when code includes a WAR dependence between operations that manipulate non-volatile variables~\cite{ratchet,dino,alpaca}.

Figure~\ref{fig:code_demo_incosistency} illustrates how state can become inconsistent in an intermittent execution using the cyclic redundancy check (CRC) code from MIBench2~\cite{hicks_mibench2_2016}. The code computes the CRC for an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder $\texttt{Rmnd}$.

Despite checkpointing at line 3 in Fig.~\ref{fig:crc_example}, the code may compute  $\texttt{data}$ incorrectly because of the read of $\texttt{Rmnd}$ on line 4 and the write of $\texttt{Rmnd}$ on line 5. Figure~\ref{fig:crc_example_war} shows an intermittent execution. The second iteration writes $\texttt{Rmnd}$, but power fails before the checkpoint. After restarting, line 9 reads the updated value of $\texttt{Rmnd}$, producing an incorrect $\texttt{data}$ value. Checkpoint-based systems risk violating memory consistency~\cite{dino}. To maintain consistency, prior systems~\cite{dino,ratchet} {\em version} a subset of non-volatile data with the checkpoint. Task-based systems~\cite{chain,alpaca}, which we focus on, ensure consistency using programmer, compiler, and runtime support.

\subsection{Task-based Intermittent Programming}
\label{section:background_task_computing}

@@

\textbf{Memory Consistency Enforcement Overhead.} Intermittent execution models incur overheads to checkpoint data~\cite{dino,ratchet,quickrecall,mementos}, manage channels~\cite{chain}, or privatize and commit data~\cite{alpaca}. Table~\ref{table:chechpoint_comparison} compares overheads for recent intermittent execution models (cf. \cite[Sec. 2.4]{alpaca}.) The mechanism responsible for overhead in each model varies, but the bulk of overhead in all models is a manipulating non-volatile memory. Checkpointing moves data to and from non-volatile memory, channel accesses manipulate non-volatile data~\cite{chain}, privatization copies from and commits to non-volatile memory~\cite{alpaca}, and idempotence solutions~\cite{ratchet} use {\em only} non-volatile memory.

\begin{figure}
\begin{subfigure}[t]{.49\columnwidth}
	\centering \includegraphics[width=\columnwidth]{figures/fram_write.eps}
	\caption{Cost of \emph{write} operation.}
\end{subfigure}%
\begin{subfigure}[t]{.49\columnwidth}
	\centering \includegraphics[width=\columnwidth]{figures/fram_read.eps}
	\caption{Cost of \emph{read} operation.}
\end{subfigure}
	\caption{Cost of accessing volatile (SRAM)/non-volatile (FRAM) memory during write and read operation.\todo{pico jole$\rightarrow$Picojoule}{Amjad}}\label{fig:framEnergy}
\end{figure}

%\begin{figure}
%\centering
	%\subfloat[Energy cost of \emph{write} operation]{\includegraphics[width=0.49\columnwidth]{figures/fram_write.eps}}
	%\subfloat[Energy cost of \emph{read} operation]{\includegraphics[width=0.49\columnwidth]{figures/fram_read.eps}}
	%\caption{Cost of accessing volatile (SRAM)/non-volatile (FRAM) memory during write (left) and read (right) operation.}
	%\label{fig:framEnergy}
%\end{figure}

To understand these non-volatile memory overheads, for the MSP430FR5969~\cite{msp430datasheet} MCU we measured the energy consumption of volatile (SRAM) and non-volatile (FRAM) memory operations. We used EDB~\cite{edb} to monitor voltage drop on a capacitor powering the device across 1600 read and write accesses to each memory type, randomizing to avoid caching. Figure~\ref{fig:framEnergy} summarizes the data, showing that a non-volatile memory access consumes around 1.5 times the energy of a volatile access in this device. In an intermittent device, energy cost corresponds to decreased run time between power failures. The cost of non-volatile memory motivates \sys, which virtualizes memory to primarily use SRAM.

@@

\textbf{Fixed-size Tasks are Inflexible and Inefficient.} If a programmer writes a task that consumes {\em more} energy than the device can buffer, the task will never complete using buffered energy. {\em preventing forward progress} by repeatedly re-executing. If a programmer writes tasks that consume far {\em less} energy than a device can buffer, the system may operate {\em inefficiently}. When a task and its successor both complete, the two tasks were interleaved by a task boundary that preserves progress and state (i.e., checkpoint or commit). However, absent a power failure, the work to preserve state was {\em unnecessary}, incurring overhead on the tasks' executions. Avoiding excessively costly, non-terminating tasks and short, high-overhead tasks is a programming challenge, given fixed hardware with a fixed energy buffering capacity. Buffer sizes from prior work vary widely from 20\,$\mu $F~\cite{rodriguez_tbcs_2015} to 0.1\,F~\cite{moo}. Sizing tasks complicates {\em porting} code from one device to another. An excessively costly task on a device with a small energy buffer may be a relatively short, task on a device with a larger buffer. The key problem is that using existing systems, the programmer statically sizes tasks for a fixed energy buffer. \sys's {\em task coalescing}, described
in Section~\ref{sec:task_coalescing}, is motivated by these observations.
